Using Deep Learning to Identify Cognitive Impairment in Electronic Health Records

### 2021-09-08 Submission/Paper/ML4H/Paper.tex
reorganized figures and tables

\paragraph{Dataset}
\label{sec:Dataset} Our gold-standard dataset consisted of a cohort (N = 16,428) of patients from the Mass General Brigham (MGB) HealthCare (formerly Partner's Healthcare, comprising two major academic hospitals, community hospitals, and community health centers in the Boston area) system who were older than 60 years (as of July 13, 2021) and had APOE genotype (biggest genetic risk factor for dementia) data available from the BioBank. Each patients' EHR record was then annotated by neurologists using a web-based annotation tool (UI Interface in Appendix A) as 1) Yes, i.e., patient has CI; 2) No i.e., Patient does not have CI; and 3) Neither i.e., sequence has no information on patient’s cognition. 
% link
\figureref{fig:fig1} shows examples of example sequences for all 3 classes.

% Figure 1 - Example Sequences
\begin{figure*}[h!]
\label{fig:fig1}
\centering 
\includegraphics[scale = 0.35]{Figure-1.png}
\caption{Example Sequences}
\end{figure*}

\paragraph{Preprocessing}
\label{sec:Preprocessing} For each patient in our gold-standard dataset, we extracted unstructured clinician notes, identified matches to dementia-related keywords (Appendix B), and constructed sequences from the note text around each of these matches (of length 800 characters). Our cohort of 16,428 patients had 279,224 sequences in total.

@@

We evaulated each model based on sequence level class assignments. Model performance for each model in shown in \tableref{tab:table2}. TF-IDF achieved AUC of 0.95 and accuracy f 0.84. The 20 words with the highest correlation coefficients using TF-IDF word vectorization are shown in Appendix \ref{app:app3}. While TF-IDF was able to identify the presence of a keyword or always pattern in a sequence, it was unable to leverage the context around each keyword match, which occasionally would negate the presence of CI. This led to the model over-zealously predicting sequences as demonstrating CI. ClinicalBERT was able to overcome these issues due to its high complexity transformer architecture, and achieved AUC of 0.98 and accuracy of 0.93. In order to generate accurate patient level class assignments, we applied ClinicalBERT to 186,730 sequences from the N = 13,941 unique patients that were not patient of our training/validation/set sets. With these sequence level predictions, we generated patient level class assignments by empirically tuning a sequence threshold that represented the number of sequences predicted as Positive by ClinicalBERT. We identified the most optimal threshold by comparing the percentage of patients being predicted as having CI (stratified by APOE allele) to the percentages of patients with CI related Meds/ICD codes (also stratified by APOE allele). \tableref{tab:table4} and \figureref{fig:fig2} shows the comparison of Med/ICD codes to ClinicalBERT patient level class assignments with a sequence threshold of 2. As shown, ClinicalBERT was able to identify a significant proportion of patients that went undetected by current clinical methods, highlighting the utility of such a tool in a clinical environment.

@@

@@

\end{table}
% \clearpage

>>>>

\paragraph{Dataset}
\label{sec:Dataset} Our gold-standard dataset consisted of a cohort (N = 16,428) of patients from the Mass General Brigham (MGB) HealthCare (formerly Partner's Healthcare, comprising two major academic hospitals, community hospitals, and community health centers in the Boston area) system who were older than 60 years (as of July 13, 2021) and had APOE genotype (biggest genetic risk factor for dementia) data available from the BioBank. Each patients' EHR record was then annotated by neurologists using a web-based annotation tool (Appendix \ref{app:slat}) as 1) Yes, i.e., patient has CI; 2) No i.e., Patient does not have CI; and 3) Neither i.e., sequence has no information on patient’s cognition. 
% link
\figureref{fig:example} in Appendix \ref{app:examples} shows examples of example sequences for all 3 classes.

\paragraph{Preprocessing}
\label{sec:Preprocessing} For each patient in our gold-standard dataset, we extracted unstructured clinician notes, identified matches to dementia-related keywords (Appendix \ref{app:keywords}), and constructed sequences from the note text around each of these matches (of length 800 characters). Our cohort of 16,428 patients had 279,224 sequences in total.

@@

We evaulated each model based on sequence level class assignments. Model performance for each model in shown in \tableref{tab:table2}. TF-IDF achieved AUC of 0.95 and accuracy of 0.84. The 20 words with the highest correlation coefficients using TF-IDF word vectorization are shown in Appendix \ref{app:app3}. While TF-IDF was able to identify the presence of a keyword or always pattern in a sequence, it was unable to leverage the context around each keyword match, which occasionally would negate the presence of CI. This led to the model over-zealously predicting sequences as demonstrating CI. ClinicalBERT was able to overcome these issues due to its high complexity transformer architecture, and achieved AUC of 0.98 and accuracy of 0.93. In order to generate accurate patient level class assignments, we applied ClinicalBERT to 186,730 sequences from the N = 13,941 unique patients that were not patient of our training/validation/set sets. With these sequence level predictions, we generated patient level class assignments by empirically tuning a sequence threshold that represented the number of sequences predicted as Positive by ClinicalBERT. We identified the most optimal threshold by comparing the percentage of patients being predicted as having CI (stratified by APOE allele) to the percentages of patients with CI related Meds/ICD codes (also stratified by APOE allele). \tableref{tab:table4} and \figureref{fig:fig2} shows the comparison of Med/ICD codes to ClinicalBERT patient level class assignments with a sequence threshold of 2. As shown, ClinicalBERT was able to identify a significant proportion of patients that went undetected by current clinical methods, highlighting the utility of such a tool in a clinical environment.

@@

% Appendix B
\section{Pictures of UI Interface for Annotations} 
\label{app:slat}
\begin{figure}[h]
\centering 
\caption{Pictures of UI Interface for Annotations}
\end{figure}

@@

\end{table}

% Figure 1 - Example Sequences
\section{Example Sequences} 
\begin{figure}[h!] \label{app:examples}
\centering 
\includegraphics[scale = 0.35]{Figure-1.png}
\caption{Example Sequences}
\end{figure}
\clearpage
